Describing demographic models is hard

### 2020-05-20 paper.tex
edits throughout

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used. The large increases in efficiency over the classical
\ms\ program~\citep{hudson2002generating} make it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is the Python API that is its primary interface, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces. In particular, programs like \ms\
required users to specify cryptic command line options to describe demographic
models \jkcomment{maybe include an example ms command line?}. Particularly for
the large models we use today, these are not intuitive or comprehensible for humans.
The Python API for \msprime\ is a great improvement, allowing the user to
state models in a documented and reproducible manner. \jkcomment{maybe show
the same model as described above in msprime notation?}

Implementing multi-population models of demographic history is still hard and error
prone, however. In this note we discuss two implementation errors that arose through
unfortunate design decisions in \msprime's demography API and which then found their way
into the scientific record. While one of these errors is not expected to have produced qualitative
mistakes, the second lead to more severe ramifications \aprcomment{departures?}. In light of these
implementation errors, we discuss improvements to \msprime's API motivated by these
discoveries and best practices for implementing and simulating complex multi-population
demography.

@@

Fortunately, the effects of this error are subtle and it would take
an exceedingly sensitive method to detect the difference.
Population sizes and structure since the time of
the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
the correct and incorrect model. However, the ancient structure distorts the distribution
of $T_{MRCA}$ beyond the split date and increases
heterozygosity in contemporary populations roughly $4\%$ over expectations
from the correct model (Fig.~\ref{fig:ooa_stats}E-F).

It is fortunate that the error in the model description has quite a
small effect because the tutorial code has been copied many times
and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repos to alert the authors.)
In most cases the publications use simulations to test some inference
method which is not directly concerned with detailed demography,
and the model is simply used as an example of a complex population
history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
\cite{zhou2018popdemog} use the incorrect model as an example
of their method for visualising demographic models can support
\msprime\ input.
% TODO ask Franz about this
Finally, \cite{pfaffelhuber2020choose}
use simulations of the incorrect model demography to evalute
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
will have any qualitative effect on their conclusions.

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
After these initial checks were made, however, the simulation
was performed without passing the parameter of demographic events,
so that the three populations remained separated with low levels of migration and
never merged as expected (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: expected heterozygosity was more than
doubled in African populations but just $30-40\%$ the expected
levels in Eurasian populations when compared to diversity under the correct model.

This simulation was performed to assess the
transferability of polygenic risk scores across human populations, in particular to
explore how human demographic history and population structure affect our ability
to predict genetic risk in populations of different ancestry than the originally studied
population.
The resulting publication has been influential in the discussion of
health inequalities and genomics, with over 350 citations since 2017.
But while the first implementation error had a small effect on resulting diversity statistics,
this error more strongly impacts simulated diversity and interpretations based on that data.
The excess divergence under the incorrect model here exaggerated the role that demography
plays in limiting the transferability of genetic risk scores across populations.

@@

Second, from a user's perspective, steps should be taken to validate the implemented
demographic model. This could include verification through code review or an
independent implementation of the model, which is the approach taken by
\stdpopsim~\citep{adrion2019community} to compile quality-controlled
models and simulation resources for a growing number of commonly studied species.
Basic statistical validation should also be performed on any simulated model, and
recent progress in computing summary statistics from tree sequence data can make
this easier~\citep{ralph2020efficiently}. If simulations are so large that statistical
validation is burdensome, subsets of the simulated data can be analyzed to ensure
general diversity patterns are sound. Such tests likely would have caught the error
in~\citet{martin2017human}.

>>>>

\sgcomment{In the race to build more realistic and useful simulations of genetic diversity, 
scientific software developers often focus on computational speed and biological realism. 
User interface and application programming interfaces (API), however, also determine 
whether software is used, and whether it is used correctly.
 }

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used. Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which make it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is the Python API that is its primary interface, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces. In particular, programs like \ms\
required users to specify cryptic command line options to describe demographic
models \jkcomment{maybe include an example ms command line?}. Particularly for
the large models we use today, these are not intuitive or comprehensible for humans.
The Python API for \msprime\ is a great improvement, allowing the user to
state models in a documented and reproducible manner. \jkcomment{maybe show
the same model as described above in msprime notation?}. \sgcomment{Other software, such as SLiM, 
have developed advanced user interfaces and even dedicated programming languages 
to facilitate the specification of various modes of evolution.\cite{}
}

Implementing multi-population models of demographic history is still hard and error
prone, however. In this note we discuss two implementation errors that arose through
unfortunate design decisions in \msprime's demography API and which then found their way
into the scientific record. The first error has relatively mild effects on genetic diversity but was 
used in many publications, while the second error was used only once but had a large impact on the 
simulation results.  
In light of these implementation errors, we discuss improvements to \msprime's API motivated by these
discoveries and more general best practices for implementing and simulating complex multi-population
demography.

@@

Fortunately, the effects of this error are subtle.
Population sizes and structure since the time of
the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
the correct and incorrect model. However, the ancient structure distorts the distribution
of early $T_{MRCAs}$. The extraneous ancient population increases increases the long-term effective 
population size, leading to heterozygosity in contemporary populations
increased by roughly $4\%$ over expectations from the correctly specified 
model (Fig.~\ref{fig:ooa_stats}E-F).

It is fortunate that the error in the model description has quite a
small effect because the tutorial code has been copied many times
and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications use simulations to test some inference
method which is not directly concerned with detailed demography,
and the model is simply used as an example of a complex population
history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
\cite{zhou2018popdemog} use the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
% TODO ask Franz about this
Finally, \cite{pfaffelhuber2020choose}
use simulations of the incorrect model demography to evalute
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
will have qualitative effect on their conclusions.

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
After these initial checks were made, however, the simulation
was performed without passing the parameter of demographic events,
so that the three populations
never merged as expected and remained separated with low levels of migration   (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: expected heterozygosity was more than
doubled in African populations but just $30-40\%$ the expected
levels in Eurasian populations when compared to diversity under the correctly specified model.

This simulation was performed to assess the
transferability of polygenic risk scores across human populations. In other words, it sought to
explore how human demographic history and population structure affect our ability
to predict genetic risk in diverse populations given the well-documented unequal representation
in medical genetic studies \cite{Popejoy}. The resulting publication has been influential in the discussion of
health inequalities and genomics, with over 350 citations since 2017.

The large excess in divergence under the incorrect model here exaggerated the role of demography
and genetic drift in limiting the transferability of genetic risk scores across populations.

@@

Second, from a user's perspective, steps should be taken to validate the implemented
demographic model. This could include verification through code review or an
independent implementation of the model, which is the approach taken by
\stdpopsim~\citep{adrion2019community} to build a library of quality-controlled
models and simulation resources for a growing number of commonly studied species. 
Such review likely would have caught the error in the msprime documentation.

Basic statistical validation should also be performed on any simulated model, and
recent progress in computing summary statistics from tree sequence data can make
this easier~\citep{ralph2020efficiently}. If simulations are so large that statistical
validation is burdensome, subsets of the simulated data can be analyzed to ensure
general diversity patterns are sound. Such tests likely would have caught the error
in~\citet{martin2017human}.
